Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-16 documents/ms.tex
updating urls and including a comment about the noprior uncertainties

\newcommand{\columbia}{1}
\newcommand{\oxford}{3}
\newcommand{\sagan}{4}
\newcommand{\uw}{5}
\newcommand{\princeton}{2}

@@

\begin{abstract}
The light curves of spotted, rotating stars often vary in a non-sinusoidal and
quasi-periodic fashion---spots move on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing
    effective inference, such as highly degenerate or high-dimensional
    parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian
    Process with a quasi-periodic covariance kernel function.
Using this highly flexible model, we are able to sample from the posterior
    Probability Density Function (PDF) of the periodic parameter,
    marginalising over the other kernel hyperparameters using a Markov Chain
    Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
\naigrain\ simulated stellar light curves, demonstrating that our results
    are more accurate than both a sine-fitting periodogram and an
    autocorrelation function method.
We also demonstrate that it works effectively on real data, by
inferring rotation periods for \nkoimcq\ \Kepler\ stars with
previously measured periods.
Because this method delivers posteriors PDFs, it will enable hierarchical
    studies involving stellar rotation, particularly those involving
    population modelling, such as inferring stellar ages, obliquities in
    exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarise our results compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
To assess the performance of the QP-GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters, for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

Initially, we also use the same priors on the hyperparameters for the KOIs
as for the simulated light curves (Table \ref{tab:priors}).
However, we found that $\ln l$ and $\ln \Gamma$ tended toward slightly different
values than the simulations.  We also found that the allowed hyperparameter
range that we used for the simulations was too large for the KOI population,
as maybe $\sim$15\% of the fits tended toward corners in the hyperparameter
space, resulting in poor period measurements.  As a result, after this initial
test, we subsequently adjusted the priors and re-fit all the KOIs.
The final priors and parameter ranges that we used are in Table \ref{tab:koipriors}.

@@

The QP-GP inference method we present in this paper produces
more accurate rotation periods than the ACF or periodogram methods.
Additionally, because it explores the posterior PDF with MCMC,
our method produces probabilistically justified uncertainties.
Unfortunately, these uncertainties still appear to be underestimated.
Of the rotation periods recovered from the simulated light curves, only 25\%
of the measured periods lie within 1$\sigma$ of the true period,
50\% lie within 2$\sigma$, and 66\% within 3$\sigma$.
The largest outlier is 114$\sigma$ away from the true value.
\tdmcomment{I assume this is for the ACF-based prior run? (should probably be mentioned)
Are these numbers similar for the logflat prior?}
% In the majority of these cases the uncertainties are underestimated due to the
% multi-modal nature of the period posterior PDFs, which makes them difficult to
% sample. \tdmcomment{Is there any evidence of this?  Doesn't sound convincing to me.
% I would think the multi-modality would only be relevant if the posterior is
% closer to the 1/2x or 2x period, but this isn't most of them.}
We attribute these underestimated uncertainties to the model choice.
Although more appropriate than a sinusoid, the GP is still only an
 effective model.
A perfect physical model of the star would produce more representative
uncertainties.

@@

The main aim of this work was to develop a probabilistic rotation period
inference method.
Probabilistic periods are necessary for hierarchical Bayesian inference,
particularly when performing population analysis.
This new GP method is capable of generating a catalogue of probabilistic
rotation periods which could, for example, reveal the period distribution
(and therefore potentially the age distribution) of stars in the Milky Way.
There is still room for improvement since the uncertainty estimates are not
yet truly representative and it is still only an `effective' model, not a
physical one.
However, it is probabilistic and provides more accurate periods than
alternative methods.
We therefore argue that the GP method is the best method for rotation period
inference currently available.

>>>>

\newcommand{\columbia}{1}
\newcommand{\oxford}{3}
\newcommand{\uw}{4}
\newcommand{\sagan}{5}
\newcommand{\princeton}{2}

@@

\begin{abstract}
The light curves of spotted, rotating stars often vary in a non-sinusoidal and
quasi-periodic fashion---spots move on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing
    effective inference, such as highly degenerate or high-dimensional
    parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian
    Process with a quasi-periodic covariance kernel function.
Using this highly flexible model, we are able to sample from the posterior
    Probability Density Function (PDF) of the periodic parameter,
    marginalising over the other kernel hyperparameters using a Markov Chain
    Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
\naigrain\ simulated stellar light curves, demonstrating that our results
    are more accurate than both a sine-fitting periodogram and an
    autocorrelation function method.
We also demonstrate that it works effectively on real data, by
inferring rotation periods for \nkoimcq\ \Kepler\ stars with
previously measured periods.
Because this method delivers posteriors PDFs, it will enable hierarchical
    studies involving stellar rotation, particularly those involving
    population modelling, such as inferring stellar ages, obliquities in
    exoplanet systems, or characterising star-planet interactions.
The code is available at \url{https://github.com/RuthAngus/GProtation/} .
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarise our results compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
To assess the performance of the QP-GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters, for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.
The posterior samples of the model parameters for the simulated light curves
are available online:
\url{https://zenodo.org/record/292340\#.WKWpiBIrJE4}.

@@

Initially, we also use the same priors on the hyperparameters for the KOIs
as for the simulated light curves (Table \ref{tab:priors}).
However, we found that $\ln l$ and $\ln \Gamma$ tended toward slightly different
values than the simulations.  We also found that the allowed hyperparameter
range that we used for the simulations was too large for the KOI population,
as maybe $\sim$15\% of the fits tended toward corners in the hyperparameter
space, resulting in poor period measurements.  As a result, after this initial
test, we subsequently adjusted the priors and re-fit all the KOIs.
The final priors and parameter ranges that we used are in Table \ref{tab:koipriors}.
The posterior samples of the model parameters for the \Kepler\ objects of
interest are available online:
\url{https://zenodo.org/record/292340\#.WKWpiBIrJE4}.

@@

The QP-GP inference method we present in this paper produces
more accurate rotation periods than the ACF or periodogram methods.
Additionally, because it explores the posterior PDF with MCMC,
our method produces probabilistically justified uncertainties.
Unfortunately, these uncertainties still appear to be underestimated.
Of the rotation periods recovered from the simulated light curves, using the
ACF-informed prior, only 25\% of the measured periods lie within 1$\sigma$ of
the true period, 50\% lie within 2$\sigma$, and 66\% within 3$\sigma$.
The largest outlier is 114$\sigma$ away from the true value.
The median uncertainty is underestimated by around a factor of two.
These numbers are similar for the uninformative prior results which also
provides underestimated uncertainties.
% In the majority of these cases the uncertainties are underestimated due to the
% multi-modal nature of the period posterior PDFs, which makes them difficult to
% sample. \tdmcomment{Is there any evidence of this?  Doesn't sound convincing to me.
% I would think the multi-modality would only be relevant if the posterior is
% closer to the 1/2x or 2x period, but this isn't most of them.}
We attribute these underestimated uncertainties to the model choice.
Although more appropriate than a sinusoid, the GP is still only an
 effective model.
A perfect physical model of the star would produce more representative
uncertainties.

@@

The main aim of this work was to develop a probabilistic rotation period
inference method.
Probabilistic periods are necessary for hierarchical Bayesian inference,
particularly when performing population analysis.
This new GP method is capable of generating a catalogue of probabilistic
rotation periods which could, for example, reveal the period distribution
(and therefore potentially the age distribution) of stars in the Milky Way.
There is still room for improvement since the uncertainty estimates are not
yet truly representative and it is still only an `effective' model, not a
physical one.
However, it is probabilistic and provides more accurate periods than
alternative methods.
We therefore argue that the GP method is the best method for rotation period
inference currently available.
The code used in this project is available at
\url{https://github.com/RuthAngus/GProtation/}.
